# Working with R - Practical {.unnumbered}

```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```

```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```

```{r load_data, context="data", include=FALSE}
library(survival)
```

## Preface {data-progressive="FALSE"}

Open Rstudio to do the practicals. Note that tasks with \* are optional.

### R packages

In this practical, a number of R packages are used. The packages used (with versions that were used to generate the solutions) are:

-   `survival` (version: `r packageVersion("survival")`)

*`r R.version.string`*

### Dataset

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html

## Basic calculations

### Expressions {.tabset .tabset-fade .tabset-pills}

::: {.panel-tabset .nav-pills}
#### Task 1

Use R as a calculator to calculate the following values:

-   $17 ^4$,
-   $45 - 2 \times 3$
-   $(45 - 2) \times 3$

#### *Hint 1*

Enter the calculations in an R-script and then run the lines using {{< kbd Ctrl-Enter >}}.

```{r eval=FALSE, include=TRUE}
17^4
2^(1/3)
```
etc

#### Solution 1

```{r save-expres, solution = TRUE, eval=F, include=TRUE}
17^4
2^(1/3)
45-2 * 3
(45-2) * 3

```


:::


### Assignment {.tabset .tabset-fade .tabset-pills}
::: panel-tabset

#### Task 2

Now use assignment to do the same calculation. Assign names `a`, `b` and `c` to each of the three values involved, then do the calculation while assigning the name `d` to the result.

#### Hint 2

Execute the following code.

```{r}
a <- 45
b <- 2
c <- 3
d <- (a - b) * c 
```

`a`, `b`, `c`, and `d` are variables. To see their values, you can just type the variable name (e.g. `a`) and hit {{< kbd Ctrl-Enter >}} or use the command `print(a)`.

#### Solution 2 {.unnumbered}

```{r}
a <- 45
b <- 2
c <- 3
d <- (a - b) * c 
print(a)
print(b)
print(c)
print(d)
d
```
:::


## Importing and Saving Data

### Save your work {.tabset .tabset-fade .tabset-pills}

It is important to save your work.

::: {.panel-tabset .nav-pills}
#### Task 1

Save the vectors `numbers <- c(34, 24, 19, 23, 16)`, `numbers_2 <- c(1:200)` and `treatment <- c("yes", "yes", "no", "no", "no", "yes")`. Use the name `new_vectors`.

#### *Hint 1*

Use the function `save(...)`. Note that you need to set the working directory.

#### Solution 1

```{r save-solution, solution = TRUE, eval=F}
numbers <- c(34, 24, 19, 23, 16)
numbers_2 <- c(1:200)
treatment <- c("yes", "yes", "no", "no", "no", "yes")
save(numbers, numbers_2, treatment, file = "new_vectors.RData")
```
:::

::: {.panel-tabset .nav-pills}
#### Task 2 \*

Save the vectors `events <- heart$event` and `eyes <- retinopathy$eye`. Use the name `vectors_survival`.

#### *Hint 2* \*

Use the function `save(...)`. Note that you need to set the working directory.

#### Solution 2 \*

```{r save2-solution, solution = TRUE, eval=F}
events <- heart$event
eyes <- retinopathy$eye
save(events, eyes, file = "vectors_survival.RData")
```
:::

### Load your work {.tabset .tabset-fade .tabset-pills}

Let's continue working on the data sets by loading our results.

::: {.panel-tabset .nav-pills}
#### Task

Load the file `new_vectors`.

#### *Hint*

Use the function `load(...)`.

#### Solution

```{r load-solution, solution = TRUE, eval=F}
load("new_vectors.RData")
```
:::

### Remove your work {.tabset .tabset-fade .tabset-pills}

Remove unnecessary objects.

::: {.panel-tabset .nav-pills}
#### Task 1

Remove the vectors `numbers`, `numbers_2` and `treatment`.

#### *Hint 1*

Use the function `rm(...)`.

#### Solution 1

```{r rm-solution, solution = TRUE, eval=F}
rm(numbers, numbers_2, treatment)
```
:::

::: {.panel-tabset .nav-pills}
#### Task 2 \*

Remove the vectors `events` and `eyes`.

#### *Hint 2* \*

Use the function `rm(...)`.

#### Solution 2 \*

```{r, eval=F}
rm(events, eyes)
```
:::
