# Data Cleaning - Practical {.unnumbered}

For this practical we will use the data set `R_data` you can find on Canvas. This is a .csv file. Load the data in R. Make sure you specify the correct file path!

```{r, message=F, echo = F, eval = T}

R_data <- read.csv("Data/R_data.csv")
```

```{r, message=F, echo = T, eval = F}

R_data <- read.csv("~/R_data.csv")
```

::: panel-tabset
### Question 1

What functions are useful for the first exploration of the data? How many observations and variables are in the data set? What type of variables are there?
:::

::: panel-tabset
### Question 2

There is a typo in one of the variable names (`Stauts` instead of `Status`). Change this.
:::

::: panel-tabset
### Question 3

Round the variable `folicacid_erys` to two decimals.

Verify by evaluating the first `20` values of this new variables (there are several ways to do this).
:::

::: panel-tabset
### Question 4

Make a new variable `birhtweight_kg` that gives the birth weight in kilo's What did you choose for the number of decimals to round the variable?

Verify by evaluating the first `10` values of this new variables (there are several ways to do this).
:::

::: panel-tabset
### Question 5

The variables `pregnancy_length_weeks` and `pregnancy_length_days` together denote the total length of the pregnancy. For example: `pregnancy_length_weeks = 38` and `pregnancy_length_days = 4`, means this patient is pregnant for 38 weeks plus 4 days. Combine the variables to obtain the length of the pregnancy in days.

Verify by evaluating the first `12` values of this new variables (there are several ways to do this).
:::

::: panel-tabset
### Question 6

Divide the variable `BMI` into categories: `<18.5` ("Underweight"), `18.5 - 24.9` ("Healthy weight"), `25 - 29.9` ("Overweight"), and `>30` (Obesity). How many patients (and %) are in each category?
:::

::: panel-tabset
### Question 7

For a current analysis, I am only interested in the patients with "Healthy weight". Additionally, I only want to look at the relation between `Status` and `birthweight`. Make a data set with only these two variables and `patientnumber`, for a subset of the data with the patients with "Healthy weight".

What are the dimensions of this data set? First try to think yourself and then check with R code.
:::

::: panel-tabset
### Question 8

Make a third data set containing the variables: `patientnumber`, `Status` and `BMI`. Then merge the data set you created in Question 7 with this data set.

Merging these two data sets can be done several different ways. Describe two ways. What are the dimensions of these data sets?
:::

::: panel-tabset
### Question 9

There are several biomarkers collected in the data set. Investigate whether there are outliers in the biomarkers: `cholesterol`, `triglycerides`, and `vitaminB12`. Which functions did you use?
:::

::: panel-tabset
### Question 10

In question 9 we found an outlier. How do you deal with this outlier?
:::

::: panel-tabset
### Question 11

Fill in the table with summary statistics below

|                             |                    | Intellectual disability | Normal brain development |
|-------------------|:-----------------|:----------------:|:----------------:|
|                             |                    |        (n = ...)        |        (n = ...)         |
| BMI, median \[IQR\]         |                    |                         |                          |
|                             | missing (n = )     |                         |                          |
| Educational level           | Low, n(%)          |                         |                          |
|                             | Intermediate, n(%) |                         |                          |
|                             | High, n(%)         |                         |                          |
|                             | missing (n = )     |                         |                          |
| Smoking                     | No, n(%)           |                         |                          |
|                             | Yes, n(%)          |                         |                          |
|                             | missing (n = )     |                         |                          |
| SAM, mean (SD)              |                    |                         |                          |
|                             | missing (n = )     |                         |                          |
| SAH, median \[IQR\]         |                    |                         |                          |
|                             | missing (n = )     |                         |                          |
| Vitamin B12, median \[IQR\] |                    |                         |                          |
|                             | missing (n = )     |                         |                          |

### Hint

You can change the order of factors

`R_data$educational_level <- factor(R_data$educational_level, levels = c('low', 'intermediate', 'high'))`
:::
