# Functions 1: Questions {.unnumbered}

## Function Arguments

::: panel-tabset
### Question 1

Look at the help file for the function `mean()`.

How many arguments does the function have?

What types of vectors are accepted?

What is the default setting for dealing with NA values?
:::

::: panel-tabset
### Question 2

Use the function `mean()` to calculate the mean of the following values:

::: callout-note
note the `NA` and use named argument matching
:::

```{r, eval = F}
c(1, 2, NA, 6)
```
:::

::: panel-tabset
### Question 3

Do Q2 again but rearrange the arguments.
:::

::: panel-tabset
### Question 4

Do Q2 again using positional matching.
:::

::: panel-tabset
### Question 5

Determine the class of `mean()` using `class()`.
:::

::: panel-tabset
### Question 6

Determine the class of `mean()` using `str()`.
:::

::: panel-tabset
### Question 7

Determine the class of the value output in Q4 using `class()`.
:::

::: panel-tabset
### Question 8

Determine the class of the value output in Q4 using str().
:::

::: panel-tabset
### Question 9

Use R as a calculator to calculate the following values:

17\^4=?

45-2\*3=?

(45-2)\*3=?
:::

## Use the operators %% and %/% to do the following:

::: panel-tabset
### Question 10

Calculate the remainder after dividing 29,079 into 184,277,809.
:::

::: panel-tabset
### Question 11

How many times does 29,079 go into 184,277,809 (i.e. what's the *integer divide* value)?
:::

::: panel-tabset
### Question 12

Do the last calculation from Q8 in another way, like this:

```{r}
a <- 45
b <- 2
c <- 3
d <- (a - b) * c
```

Now check what a, b, c, and d are. You can just type the variable name (e.g. a) and hit 'Control+Enter' or use the command `print(a)`.
:::

## if(), else, and ifelse() and Vectorization

::: panel-tabset
### Question 13

Write a function called 'evenOrOdd' involving `if` and `else` that takes an argument `x` and returns "Even" or "Odd" depending on whether or not `x` is divisible by 2. (Do not use the ifelse() function).
:::

::: panel-tabset
### Question 14

Is your function 'evenOrOdd' vectorized? Check by passing it the vector:

```{r}
w <- c(3, 6, 6, 4, 7, 9, 11, 6)
```
:::

::: panel-tabset
### Question 15

Another way to determine if each element of a vector is even or odd is to use the `ifelse()` function, which serves as a vectorized version `if` and `else`. Use `ifelse()` to obtain "Even" or "Odd" for each element of `w`.
:::

## Logical Operators &, \|, and !

::: panel-tabset
### Question 16

What will be the result of the following:

```{r, eval=F}
(10 < 20 && 15 < 16) || 9 == 10
```
:::

::: panel-tabset
### Question 17

One of the following evaluates to TRUE, the other to FALSE. Which is which?

```{r, eval=F}
4 < 3 && (5 < 6 || 8 < 9)
```

```{r,eval=F}
(4 < 3 && 5 < 6) || 8 < 9
```
:::

::: panel-tabset
### Question 18

The data set in 'BPressure.txt' contains the systolic and diastolic blood pressure readings for 22 patients.

a)  Read the data from BPressure.txt into a data frame called bp using read.table().

b)  A person's blood pressure is classified as normal if the systolic level is below 120 and the diastolic level is below 80. Use relational, and logical operations in square brackets `[ ]` to extract from bp the rows corresponding to patients with normal blood pressures.

c)  Now use the same strategy to extract the rows corresponding to patients whose blood pressures aren't normal.
:::
