# Creating your own functions {.unnumbered}

To write your own function, you need to use the function `function()`. In the brackets `()` you specify argument(s) that will be used in the function (with our without defaults), and in curly brackets `{ }` you specify what the function should do, referring to the argument(s).

The general form is:

```{r}
myFun <- function(arg1, arg2) {
  ## Here you type expressions that use the arguments
}
```

Each line inside the function is an object assignment, a function call, a subsetting, a conditional statement, an if/else statement, a for loop, etc. - basically anything you have now learned how to do in R that you want the function to do!

Below is an easy example that calculates the mean of two values (`x` and `y`):

```{r}
mean_xy <- function(x, y){
  (x + y)/2
}
```

This function can be used the same way as how you have been using functions before:

```{r}
mean_xy(2,6)
```

Or:

```{r}
mean_xy(x = 2, y = 6)
```

To have a function output something, you must return something. Either the value of the last command is returned (as in `mean_xy`) or you can use `return()`.

```{r}
mean_xy <- function(x, y){
  z <- (x + y)/2
  return(z)
}

mean_xy(x = 2, y = 6)
```

Here are a few other examples.

```{r}
mean_xy_2 <- function(x, y){
  z <- (x + y)/2
  x
  z
}

mean_xy_2(x = 1, y = 3)
```

Note that x is not returned. Only the last expression is returned.

```{r}
mean_xy_3 <- function(x, y){
  z <- x + y 
  return(x)
  z
}

mean_xy_3(x = 1, y = 3)
```

Note that z is not returned, if a return statement is encountered in the function anything after that statement is [not executed]{.underline}.

You can create functions with a variable number of arguments using `...`. For example, here’s a function that returns the mean of all the values in a vector of arbitrary length:

```{r}
mean_vector <- function(...){ 
  z <- mean(c(...))
  return(z)
} 

mean_vector(1,2,3)
mean_vector(1,2,3,4,5,6,7,8,9,10) 
```

The arguments in a vector do not have to be single values. Functions can be vectorized:

```{r}
mean_x <- function(x){ 
  z <- mean(x)
  return(z)
} 
x <- c(1,2,3,4,5)
mean_x(x)

```

Custom functions in R are useful if you have a bunch of commands that you have to use multiple times. By combining them in a function you 1) save time, 2) keep your code concise, and 3) make less coding mistakes.

In the next example a function called `my_descriptives` is made to calculate a mean of a vector only for the positive values.

```{r}
my_descriptives <- function(x){
  x.trim <- x[x>=0]
  out <- mean(x.trim)
  return(out)
}
```

In the first line inside the function a sub sample of the vector is taken `x.trim`, with only values `>=0`. In the second line, the mean of this `x.trim` is taken.

This function can be used to describe a vector in my data set, but there are negative values where only positive values are allowed.

```{r, echo=FALSE}
PatientID <- 1:25
Ages <- round(c(rnorm(15, mean = 45, sd = 10), -50, rnorm(9, mean = 45, sd = 10) ) )
data <- data.frame(PatientID = PatientID, Ages = Ages)
data$Sex <- as.factor(c(rbinom(24, 1, 0.5),-1))
```

In this data set, there is a variable `Ages`:

```{r}
data$Ages
```

There is one value `-50`, that is clearly an error.

```{r}
my_descriptives(data$Ages)
```

Compare the output with using the standard function `mean()`:

```{r}

mean(data$Ages)
```

In the standard `mean` function, the negative outlier is included and influences the mean!

The output of a function does not need to be a scalar. This version of the function `my_decriptives()` provides the whole summary of the variable, instead of only the mean.

```{r}
my_descriptives <- function(x){
  x.trim <- x[x>=0]
  out <- summary(x.trim)
  return(out)
}

```

```{r}
my_descriptives(data$Ages)

```

Again, let's compare the output to the standard `summary()` function.

```{r}
summary(data$Ages)
```

If you have multiple objects to return, you have to put them in an object container, like a list, vector, array or data.frame. It is not possible to return multiple individual objects like this:

`return(x,y)`

but it is possible to return them in a vector or list like this:

`return(c(x,y))`

`return(list(x,y)`

Here is an example of the function with multiple outputs:

```{r}
my_descriptives2 <- function(x){
  x.trim <- x[x>0]
  below0 <- sum(x<0)
  meanX <- mean(x.trim)
  return(list(below0, meanX))
}
```

The function additionally returns how many values were negative.

```{r}
my_descriptives2(data$Ages)
```

There was `1` value below zero, as provided by the first element in the list.

Notice how the function gives an error if you do not put the items in a list:

```{r, error=TRUE}
my_descriptives2_wrong <- function(x){
  x.trim <- x[x>0]
  below0 <- sum(x<0)
  meanX <- mean(x.trim)
  return(below0, meanX)
}

my_descriptives2_wrong(data$Ages)
```

Specifying default arguments of a function, can be done by filling in the default value in the `function()` call. Here is an example of a function with a default argument (`y = 2`).

```{r}
calc4 <- function(x, y = 2){ 
  z1 <- x + y
  z2 <- x * y 
  return(c(z1, z2))
} 

calc4(x = 1) ## uses y = 2
calc4(x = 1, y = 3) ## overwrites default value of y
```

## Function environments

Each function, whether built-in or user-defined, has an associated environment, which can be thought of as a container that holds all of the objects present at the time the function is created.

When a function is created on the command line, it's environment is the so-called "Global Environment":

```{r}
w <- 2
f <- function(y) {
  d <- 3
  return(d * (w + y))
}
environment(f)
```

The function `objects()` (or `ls()`), when called from the command line, lists the objects in the Global Environment:

```{r}
objects()
```

### Global and Local Variables

In the function `f()` defined above, the variable `w` is said to be global to `f()` and the variable `d`, because it's created within `f()`, is said to be local to `f()`. Global variables (like `w`) are visible from within a function, but local variables (like `d`) aren't visible from outside the function. In fact, local variables are temporary, and disappear when the function call is completed:

```{r, eval=FALSE}
f(y = 1)
d
```

You get an error: [Error in eval(expr, envir, enclos) : object 'd' not found]{style="color: red"}, indicating that the variable d does not exist in the 'Global Environment'.

When a global and local variable share the same name, the local variable is used:

```{r}
w <- 2
d <- 4

f <- function(y) {
  d <- 3
  return(d * (w + y))
}

f(y = 1)
```

Note also that when an assignment takes place within a function, and the local variable shares its name with an existing global variable, only the local variable is affected:

```{r}
w <- 2
d <- 4 # This value of d will remain unchanged.

f <- function(y) {
  d <- 3 # This doesnt affect the value of d in the global environment
  return(d * (w + y))
}

f(y = 1)

d
```
