## Markdown: Hints {.unnumbered}

### Question 1

When using the template Rstudio this should not give any problems. Remember that the markdown code to start a section consists of two hashes `##`.


### Question 2

To code chunk you should insert in the Setup section should look something like this:

```{r setup}
dat <- read.csv("Data/R_data2.csv")
```

### Question 3

You can cary out all the transformationns using code like:

  
```{r transformations}
dat$pregnancy_length <-  dat$pregnancy_length_weeks *7 +
    dat$pregnancy_length_days

dat$BMI_cat <- cut(dat$BMI, breaks=c(-Inf, 18.5, 24.9, 29.9, Inf),
                       labels=c("Underweight", "Healthy weight", "Overweight", "Obesity"))

dat$log_homocysteine <- log(dat$homocysteine)

dat$log_vitaminB12 <- log(dat$vitaminB12)

for(i in 1:length(names(dat))){
  if(class(dat[[i]]) == "character"){
    dat[[i]] <- as.factor(dat[[i]])
  }
}

dat$Status <- factor(dat$Status, 
                           levels = c("normal brain development", "intellectual disability"))


dat <- dat[ , !(names(dat) %in% c("pregnancy_length_weeks", "pregnancy_length_days", "BMI", "homocysteine", "vitaminB12"))]
```      

### Question 4

To all the desciptives for all variables in the data set you can use the following code template:

```{r}
for(i in 1:length(names(dat))){
  if(class(dat[[i]]) == "numeric"){
    # do something for numeric vriables
  } else if(class(dat[[i]]) == "factor"){
    # do something for factor variables
  }
}

```


Histograms can be created in a simular loop. To make the actual plot use the `hist` function.

### Question 5

Use the `t.test` or `wilcox.test` function to compare the mean of the logarithm of the Vitamin B12 for the two levels of `Status` (normal brain development or intellectual disability).

### Question 6

Use the `glm` function to perform logistic regression analysis to investigate the association between `Status` and log `Vitamin B12` while adjusting for `medication`, `smoking` and `alcohol`.

The code should look something like:

```{r eval=FALSE}
glm1_adjusted <- glm(formula, data = dat, family = binomial)

```

Use the `summary`, `coef` and `confint` functions to extract the following details.



### Question 7

State your conclusions from the adjusted and unadjusted analysis.