# Conditional execution using if and else

### If? Then! Else?

Sometimes we'll want R to execute a statement only if a certain condition is met. This can be accomplished via the `if()` and (optionally) `else` statements:

```{r, eval=FALSE}
if()    # Used to execute a statement only if the given condition
        # is met
else    # Used to specify an alternative statement to be executed 
        # if the condition given in if() isn't met
```

Such ***conditional execution*** commands have the forms:

```{r, eval=FALSE}
if (condition) {
  statement1
}
```

and

```{r, eval=FALSE}
if (condition) {
  statement1
} else {
  statement2
}
```

where `condition` is a logical expression (i.e. it evaluates to `TRUE` or `FALSE`)

In both cases above, if `condition` is `TRUE`, `statement` is executed. If `condition` is `FALSE`, then in the first case nothing happens, but in the second case, `statement2` is executed.

Here's a simple example:

```{r}
x <- 5
if (x < 10) {
  y <- 0
}
y
```

Here's another:

```{r}
if (x >= 10) {
  y <- 1
  } else {
    y <- 0
}
y
```

It is also possible to write short if/else statements on a single line, in that case you do not have to include `{ }`:

```{r}
y <- if(x >= 10) 1 else 0
y
```

Be aware that when using such ***conditional assignment*** statements, in the absence of `else`, `if()` returns `NULL` if the condition isn't met. So:

```{r}
y <- if(x >= 10) 1
y
```

In the next example, `return()` is used to terminate a function call and return a value that depends on whether or not a condition is met:

```{r}
mySign <- function(x) {
    if(x < 0) {
      return("Negative")
    } else {
      return("Non-negative")
    }
}
```

We get:

```{r}
mySign(13)
```

### Nested `if()` Statements

In addition, `if()` statements can be nested. The form for nested `if()` statements is

```{r, eval=FALSE}

if(condition1) {
  if(condition2) {
    statement1 
  } else {
    statement2
  }
}
```

The `else` always refers to the most recent `if()`, but to keep our code readable, we use tab indentation for every level of nesting in our nested if-else statements.

Apart form nesting `if()` statements it is also possible to string multiple if statements together like so:

```{r}
whatAnimalSound <- function(animal){
  if(animal == "cat") {
    return("Meow!")
  } else if (animal == "frog") {
    return("Ribbit!")
  } else if (animal == "dog") {
    return("Woof!")
  } else {
    return(paste0("I don't know what sound a '",animal,"' makes..."))
  }
}
```

```{r}
whatAnimalSound("dog")
whatAnimalSound("bird")
```

::: callout-tip
Notice that the last statement can be `else`, but to string together multiple statements you have to use `if else`.
:::

### Vectorized if-else: The `ifelse()` Function

Sometimes we'll need to create a vector whose values depend on whether or not the values in another vector satisfies some condition. In that case we can use the `ifelse()` function, which works on a vector of values by repeating the same conditional statement for every value in the vector.

`ifelse()` takes argument `test`, the condition to be met, `yes`, the return value (or vector of values) when `test` is `TRUE`, and `no`, the return values (or vector of values) when `test` is `FALSE`.

For example, here we convert the values in `height` to `"short"` or `"tall"` based on whether they are larger than 69 or not:

```{r}
height <- c(69, 71, 67, 66, 72, 71, 61, 65, 73, 70, 68, 74)

htCategory <- ifelse(height > 69, yes = "tall", no = "short")

htCategory
```

The `ifelse` function is a very simple way of applying a test to a vector of values. To apply a more complicated function to a vector or to apply a function to multiple rows of a matrix or a dataframe we can use the `apply` functions which will be discussed later on.
