---
title: "Solution Rmarkdown"
author: "Sten Willemsen"
date: "2024-06-06"
output: 
  html_document:
    code_folding: hide
---

## Setup

```{r setup}
dat <- read.csv("Data/R_data2.csv")
```


## Introduction

## Data

We perform the following data transformation steps:

* We define the variable `pregnancy_length` by adding together. 

*  We define the variable `BMI_cat` by dividing the variable BMI into categories: <18.5 ("Underweight"), 18.5 - 24.9 ("Healthy weight"), 25 - 29.9 ("Overweight"), and >30 (Obesity).

* We log transform `homocysteine` and `vitaminB12`.

* We transform the categorical variables to factors. 

* We remove the original variables `pregnancy_length_weeks`, `pregnancy_length_days`,  `BMI` and `homocysteine` and `vitaminB12` from the data set.

```{r transformations, echo=FALSE, message=FALSE}
dat$pregnancy_length <-  dat$pregnancy_length_weeks *7 +
    dat$pregnancy_length_days

dat$BMI_cat <- cut(dat$BMI, breaks=c(-Inf, 18.5, 24.9, 29.9, Inf),
                       labels=c("Underweight", "Healthy weight", "Overweight", "Obesity"))

dat$log_homocysteine <- log(dat$homocysteine)

dat$log_vitaminB12 <- log(dat$vitaminB12)

for(i in 1:length(names(dat))){
  if(class(dat[[i]]) == "character"){
    dat[[i]] <- as.factor(dat[[i]])
  }
}

dat$Status <- factor(dat$Status, 
                           levels = c("normal brain development", "intellectual disability"))


dat <- dat[ , !(names(dat) %in% c("pregnancy_length_weeks", "pregnancy_length_days", "BMI", "homocysteine", "vitaminB12"))]
```



## Analysis and Results 

We show descriptives of all variables in the data set.


### Descriptives

```{r descriptives}
for(i in 1:length(names(dat))){
  if(class(dat[[i]]) == "numeric"){
    print(paste("Mean of", names(dat)[i], ":", mean(dat[[i]])))
    print(paste("Standard deviation of", names(dat)[i], ":", sd(dat[[i]])))
  } else if(class(dat[[i]]) == "factor"){
    print(paste("Frequency of", names(dat)[i], ":"))
    print(table(dat[[i]]))
  }
}

```
For the continuous ones we also make a histogram.

```{r}
for(i in 1:length(names(dat))){
  if(class(dat[[i]]) == "numeric"){
    hist(dat[[i]], main = paste("Histogram of", names(dat)[i]))
  }
}
```

### Unadjusted Analysis

We compare the mean of the logarithm of the Vitamin B12 for the two levels of `Status` (normal brain development or intellectual disability).

```{r}
t.test(log_vitaminB12 ~ Status,data = dat)
```

## Adjusted analysis

We now perform logistic regression analysis to investigate the association between `Status` and log `Vitamin B12` while adjusting for `medication`, `smoking` and `alcohol`.

```{r}
glm1_adjusted <- glm(Status ~ log_vitaminB12 +  medication + smoking + alcohol, data = dat, family = binomial)

summary(glm1_adjusted)
coef(glm1_adjusted)
confint(glm1_adjusted)

```




## Conclusion and Discussion

**Main points:**

* In the unadjusted analysis, we could not show that the mean of the logarithm of the Vitamin B12 is significantly different for the two levels of `Status`.

* In the adjusted analysis, we found that the log `Vitamin B12` is not significantly associated with `Status` while adjusting for `medication`, `smoking` and `alcohol`.

